
# Getting Started with R {-}
<hr>

## Why use R? {-}

R is a statistical software environment that is widely used by statisticians, social scientists, and data analysts. R is different from "point-and-click" software packages like Microsoft Excel, SPSS, or Tableau in that it requires the user to write code via a command line interface. For this reason, R is also often referred to as a programming language. However, the R environment is much more interactive than other programming languages like C or Java which makes it easier to learn and use.

Here are a few key advantages of R.

- R is **free** and **open source**^[https://opensource.org] 
- R is available on Windows, Mac OS, and UNIX/Linux
- R is **flexible**: you can write, modify, save, and share your own code
- R is **powerful**: you can do everything from making [high-quality graphics](https://www.r-graph-gallery.com) to running [sophisticated statistical machine learning models](https://www.r-bloggers.com/101-machine-learning-algorithms-for-data-science-with-cheat-sheets/)
- R is **popular**: there is a large and growing online community of users making it easy to find answers to any problem you run into

Lastly, learning R is a tangible and highly-valued skill you can put on your CV!

## Downloading R and RStudio {-}

To download R, go to http://cran.r-project.org/mirrors.html and choose the CRAN mirror closest to your physical location. Then select "Download R for..." and choose your operating system. For other questions, see the [R FAQ](https://cran.r-project.org/doc/FAQ/R-FAQ.html).

After downloading R, you should also download [RStudio](https://rstudio.com), which is an integrated development enviroment (IDE) for R. In short, it provides a much more interactive and user-friendly interface for using R. To download, go [here](https://rstudio.com/products/rstudio/download/?utm_source=downloadrstudio&utm_medium=Site&utm_campaign=home-hero-cta),  select "RStudio Desktop - Free", and select the installer corresponding to your operating system.

## Installing and Loading Packages {-}

As you will see, one of the most attractive features of R is its library of over 10,000 packages. R packages -- which are collections of R functions, code, and data sets -- allow us to use code written by others in order to use certain data sets, make certain graphs, or run certain models. 

For example, in this course we will discuss a variety of regression models including linear regression models and regression trees. While the R function to estimate a linear regression model (called  `lm`) is included in "base" R, the function to estimate a regression tree is not. Rather than writing the code ourselves, we can download an R package! 

One package that provides code for estimating regression trees is called `rpart`. To use functions within the `rpart` package, we must first install it.

```{r eval = FALSE}
install.packages("rpart")
```

Alternatively, you can navigate to the "Packages" tab in RStudio (likely in the lower right panel), click "Install", and search for **rpart**.

**Note: You only need to install a package once!** After a package is installed, it will remain installed until you upgrade your version of R/RStudio.

However, in each R session (i.e., each time you open RStudio), you will need to *load the package*.

```{r eval = FALSE}
library("rpart")
```

Again, an alternative is to navigate to the "Packages" tab in RStudio, find the package name, and click on the box to the left of the name.
